Master's Research Proposal (DRAFT)

### 2024-09-26 proposal/Proposal.tex
edits

@@

\textbf{Target:} Depth range. \newline

@@

\textbf{Target:} Spatial grid cell selection as movement. Any state feature updates needed to capture internal states of a cohort. \newline

@@

\begin{enumerate}
\item Formatting Data for Odds Modeling 
\item Class Balancing
\item Splitting Data into Train, Validation, Test
\item Training and Hyperparameter Tuning Using Standard Model Types
\item Diagnoses of Fit and Comparison of Different Model Types
\item Identifying Sample Groups that are More or Less Fit
\end{enumerate}

@@

\begin{figure}[h!] 
  \includegraphics[width=\linewidth]{simulation.png}
  \caption{Simulation Outline}
  \medskip
	\small
	Boxes with rounded edges would be inputs specified by the user. The hexagons are functions 		provided by the simulation and the trapezoids are data generated during the main loop. 
  \label{fig:simulation}
\end{figure}

@@

\begin{enumerate}
\item \textbf{Understanding Connectivity:} by looking at the vector flows of fish throughout the space we will be able to identify which areas are highly connected and which are more or less separated by environment, cohort, or other factors of fish behavior.
\item \textbf{Understanding Vulnerability:} by especially looking at changes in depth occupancy and preference for habitat we will attempt to quantify the degree of vulnerability to different kinds of fishing gear as a function of space and time. 
\item \textbf{Cumulative Mortality:} by being able to follow cohorts through space and time we should be able to show how different fishing pressure scenarios affect the cumulative fishing mortality.
\item \textbf{Protected Areas/Times:} Of those fishing pressure scenarios, one of especial interest is modeling Marine Protected Areas (MPAs) or other kinds of closures in space and time.
\item \textbf{Uncertainty:} we will show that such a simulation can be run at scale, bootstrapping can be applied, and therefore uncertainty in simulation outcomes quantified. 
\end{enumerate}

@@

From studies that started with Beverton and Holt's work on recruitment in fisheries, we know, with a reasonable degree of certainty, that as a general rule fish stocks operate according to an asymptotic relationship between spawning biomass and recruits \citep{waltersmartell} (Fig. \ref{fig:recruitment}). What we see is a clear pattern in recruitment verses spawners for low spawners and then eventually we reach a point where this curve asymptotes and large changes to the number of spawners creates little to no change (especially when considering the actual variability in this data) in the number of recruits. Obviously, given the real variability in nature, this pattern looks more like a scatter shot of points but plenty of studies over several species have shown this overall pattern in the mean borne out \citep{waltersmartell}. \newline

@@

The implication for fisheries management is pretty clear - sustainable fishing is in large part driven by just maintaining the spawning stock biomass (SSB) above that knee in the curve \citep{waltersmartell} \citep{king}. Indeed this is well noted as the basis of why fisheries \textit{can} exist in the first place. If taking spawners always led to decreases in recruits, we'd quickly fish a stock out of existence. It is only through the compensation we get on the asymptote that sustainable fisheries are possible. \newline

However, two very simple observations of fish biology will show how actually managing SSB can be made difficult.

The first, and simplest, is age structure. Obviously we're talking here about maintaining a specific level of spawning stock biomass. But today's juveniles are tomorrow's SSB and as fish grow they typically become more fecund. This means that in order to maintain our SSB over time we have to be careful about how we target each age group. This gets even more complicated for fish like Pacific salmon (\textit{Oncorhynchus}) or many species of groupers (\textit{Epinephelinae}) where life history is intimately entwined with spawning. The salmon obviously only breed once \citep{NOAA24} and some species of groupers are protogynous hermaphrodites \citep{RedGrouper} meaning that if you target a specific age group you are effectively targeting a specific sex. Clearly understanding and managing our age class selectivity matters when it comes to maintaining a healthy SSB for years to come. \newline

The other observation is that many species of fishes actually form multiple sub stocks within what we may consider a single fishery (or sometimes many fisheries distributed across different countries) \citep{lorenzen2010} \citep{salmonplan} \citep{prince2010}. Pacific salmon are once again a good example in that they home to specific streams and only breed with other salmon in those streams and many other species of fish form fish spawning aggregations as well \citep{erisman2017}. Research also indicates that even fish like Atlantic herring likely form distinct genetic groups due to differences in homing during breeding and school behavior \citep{armstrong2001}.

@@

All of this to say that understanding heterogeneity in vulnerability by things like age class and sub stock is of the utmost importance for managing a fishery and truly ensuring an appropriate amount of SSB remains year over year. It is also worth noting that this knowledge around changes in selectivity in space and time is not only useful for considerations around SSB but also for other management issues such as bycatch avoidance and growth overfishing.

What kind of tooling do we need in order to address these issues and test out ideas for "selectivity aware" fisheries management? Clearly we need some kind of reliable simulation. \newline

@@

Clearly then this kind of data suites our modeling and simulation purposes well as we get behavior outcomes directly.

@@

Fisheries management is as diverse a subject as the species and cultures to which it is applied \citep{waltersmartell} \citep{king}. It simultaneously has to deal with creatures from Bluefin Tuna (\textit{Thunnus thynnus}) who make large oceanic migrations around the globe to various clams whose entire life cycles can be summarized within a few square meters \citep{prince2010}. There are species that school and species that consume those schools. Some live entirely at sea, some in inland rivers and lakes, and some, like salmon and eels spend time in both. There are species whose lives last the span of a year to others that can live longer than any us human beings.

And to capture this vast array of diversity are similarly diverse strategies and gears. From hooks to gillnets, from small boats to at sea factories, from recreational fishers to roving commercial bands, and everything in between the variety in strategies that exist to catch fish are numerous \citep{waltersmartell} \citep{king}.

@@

First, it is clear that engagement is best facilitated by presentations of knowledge or data that are visual and relevant to the ways in which people normally think about their world \citep{prince2010} \citep{lorenzen2010}. By having simulations that can simply be shown and laid down on a map that places the spatio-temporal modeling in familiar context, one should be able to make objections or confirmation much easier. Rather than having to attempt to condense down loads of parameters, summary statistics, and the like in a model, one can simply show a map and folks can point at things that line up or are in disagreement with their intuition and experience. In this way mapped out views of the simulation work to be a useful device for consensus building among multiple groups of stakeholders. Not everyone may be familiar with the nuances of a specific kind of model but everyone in the problem can point to a map and understand what's going on. 
\newline

@@

For example, suppose that we have a more standard model that has been hand crafted by a scientist familiar with the species in question. They may have included specific reactions to, say, oceanic currents, temperatures, and primary productivity. But then if someone looking at the simulation points out that they think the species' behavior is also being driven by day/night patterns the scientist then has to go back to the drawing board, figure out how to add this into their overall model, code it up, tune parameters, and then return - all without any real guarantee that whoever provided the feedback is right. And this would be for one piece of feedback. If there is a lot of feedback this only gets more difficult.

@@

Finally it is worth noting that the potential in combining this modeling with fisheries dependent information is quite interesting on its own. Any forecasting model is always going to have issues predicting farther and farther into the future but fisheries dependent data could provide a means for making corrections to keep the model in line. This would amount to something like weather forecast where for the next few days the forecast is accurate enough to precisely predict specific kinds of movement helpful to fishers themselves.

@@

Fisheries assessments are one way in which unknown parameters important to management are estimated. By and large they are comprised of two models - (1) a state model that represents the underlying stock and contains such things as growth, age composition, recruitment, abundance, and so on and (2) an observation model that represents how data such as catch rates, fishery independent indices, and so on arise from the state model \citep{king}. Depending on how the assessment is being done the parameters in the state model will either be estimated independently or left as parameters in the model to be fit. The observation model on the other hand is encoded as a set of distributions dependent on those parameters (and including some parameters of its own) thus allowing for a likelihood to be computed against known data for any given set of parameters \citep{sippel2014}. This, in turn, allows for maximum likelihood estimation (MLE) routines to then fit the likely parameters of the assessment. \newline

@@

In short the pattern is (1) design a sim that has your unknown statistics as parameters, (2) specify how data is drawn from the sim, (3) fit your parameters using MLE. Then those desired statistics are extracted from the assessment (usually with some notion of uncertainty attached to them) and used to inform management.

@@

As has been noted by several authors, most standard assessments have typically assumed in their state model (or simulation) that the stock they are working with is bounded by the area under assessment, is well mixed, and constitutes a single unit \citep{sippel2014} \citep{punt2019} \citep{goethel2023}. However from studies of fish behavior and the fact that fish obviously don't obey national or state defined boundaries, these assumptions are rarely true and indeed simulation studies have shown that assuming them can lead to significant errors and biases \citep{lorenzen2010}.

@@

Here then is another potential application of the kind of model proposed in this document. As we mentioned earlier these assessments are trying to model at least three things - recruitment, local population dynamics parameters such as growth or natural mortality, and then movement. By having a movement simulation based on real observations of the species in question we would be able to help build a high quality simulation of at least the last of those three. Rather than having to assume one form of a movement or another we'd instead be able to see whether the assessment paradigm in question can capture the relevant aspects of structure in the \textit{real} movement of our stocks and sub-stocks. And this in turn would help in selecting the assessments best suited to the job (as well as illustrate what uncertainties exist if a lack of specific kinds of data requires us to use an over-simplified assessment paradigm).

@@

\begin{enumerate}
\item \textbf{High Precision as a Requirement} - IBMs typically predict what an animal \textit{will} do as opposed to the likelihoods of what it \textit{could} do. This creates the requirement for capturing \textit{all} the meaningful drivers of variability in the behavior of the animal. To see why this is consider the example of a creature that can either choose to stay put or move to another location. If the more likely of the two actions is to stay put a deterministic model \textit{must} predict that the animal will \textit{always} stay put. But this is obviously non-sensical. However because the model is deterministic in nature the modeler cannot depend on stochasticity to save the day - instead they must find whatever is driving the difference between staying put and moving on. And this can cascade into greater and greater requirements for modeling the "internal state" of an animal. 
\item \textbf{Obscurity} - given the inherent complexity of IBMs that arises from the point above it gets increasingly difficult to make IBMs clear to new users or potential developers. There can often be many dozens of different parameters, functions, and "internal states" to keep track of without even considering all of the external modeling that has to go in place to provide for all of this. And this makes for not just complicated descriptions but equally if not more complicated code.
\item \textbf{Hand Made} - almost all IBMs have been hand-coded in the sense that every functional relationship is specifically crafted and outlined by the modeler. This is obviously a very laborious process but it also means that no two IBMs are the same and that lack of generalization or abstraction makes exchanging lessons learned near difficult to impossible. 
\item \textbf{No Generalization of Fitting or Diagnostics} - because each IBM is effectively unique the methods for fitting, the data used, and the diagnostics developed are all unique as well. This once again increases the workload of the modeler, makes exchanging lessons learned and best practices extremely difficult, and also makes starting a new model from scratch quite prohibitive. 
\end{enumerate}

@@

The immature fish then spend time (up to five years) growing in the marine environment before maturing and then migrating back to the river where they were born to breed and subsequently die \citep{oceanchinook}. The timing of such homing behavior is clearly season dependent with species further north having to avoid ice build up during the cooler months \citep{oceanchinook}. The mechanism behind this homing ability is still a question under active research \citep{bracis2012}. \newline

Given this homing behavior, Chinook salmon populations are in fact a series of reasonably separated genetic sub-populations each of which can have its own life history strategies. Indeed using tagging data it has been demonstrated that both the initial and later marine distributions of Chinook salmon is highly dependent on the region of origin for the fish in question \citep{shelton2019} \citep{tucker2019}.

@@

The initial purpose of these hatcheries programs was to supplement losses in juveniles due to targeting of adult fish in river systems, however the programs have not created the kind of recovery that was intended and it is unclear how wild and hatchery populations interact. For example it is well known that there is quite a bit of genetic exchange between wild and hatchery types and concern that hatchery types are less suited to wild environments \citep{oceanchinook}. \newline

There are valuable tribal, commercial, and recreational fisheries of Chinook salmon throughout the Pacific \citep{salmonplan}.

Given the highly mobile nature of Chinook salmon independent management by sovereign nations is simply not possible and so management is the ultimate responsibility of the Pacific Fishery Management Council. Federal management councils (NOAA for the United States) then collaborate with the Pacific Fishery Management Council to set harvest quotas and other management objectives. The governments of Cananda and the United States then work with states, tribes, and industry to create and manage shared objectives based on these overall goals \citep{NOAA22}.

@@

\section{Appendices}

\subsection{Value of Log Likelihood Maximization}

\begin{theorem}
Suppose we have a set of possible outcomes $V=\lbrace v_k \rbrace$ with probabilities $P_k$. Such that:

$$\sum_k P_k = P$$

Next suppose we concoct a new series of probabilities $U_k = P_k \alpha_k$ s.t.

$$\sum_k U_k = P-\epsilon$$

where $\epsilon \geq 0$. There does not exist a set of $\alpha_k$ s.t.

$$\prod_k \left(\frac{P_k\alpha_k}{P_k}\right)^{P_kN}>1$$

\end{theorem}

\begin{proof}

We proceed by induction. \newline

First note that from the equation above we get:

$$\sum_k P_k \ln \alpha_k > 0$$

and subtracting $\sum_k U_k$ we arrive at:

$$\sum_k P_k \left( \ln \alpha_k - \alpha_k \right) > -P+\epsilon$$

Now suppose for $k$ outcomes we know no such $\alpha_k$ exist as to satisfy the above. Suppose we incorporate a new $k+1$ term s.t:

$$P_{k+1} + \sum_k {P_k} = P_{k+1} + P$$

and:

$$P_{k+1}\alpha_{k+1} + \sum_k {P_k}\alpha_k = P_{k+1} + P - \epsilon = P_{k+1}\alpha_{k+1} + \left(P - P_{k+1}(\alpha_{k+1}-1) \right) - \epsilon$$

Now we need:

$$P_{k+1}\left(\ln \alpha_{k+1} - \alpha_{k+1}\right) + \sum_k P_k \left( \ln \alpha_k - \alpha_k \right) > -P_{k+1}-P+\epsilon$$

However given our inductive assumption we know that at best,

$$\sum_k P_k \left( \ln \alpha_k - \alpha_k \right)=-P$$

therefore at best:

$$P_{k+1}\left(\ln \alpha_{k+1} - \alpha_{k+1}\right) - P > -P - P_{k+1} + \epsilon$$

Furthermore the easiest case for us is if $\epsilon=0$. If it cannot be satisfied than this certainly doesn't work for $\epsilon>0$. So let's consider:

$$P_{k+1}\left(\ln \alpha_{k+1} - \alpha_{k+1}\right) - P > -P - P_{k+1}$$

or equivalently:

$$\ln \alpha_{k+1}> \alpha_{k+1}-1$$

Now if $\alpha_{k+1}=1$ we have:

$$\ln 1 = 1-1$$

Further consider the derivatives of each side:

$$\partial_{\alpha}\ln \alpha_{k+1}=\frac{1}{\alpha_{k+1}}$$

$$\partial_{\alpha}(\alpha_{k+1}-1) = 1$$

If $\alpha_{k+1}> 1$ then the log component rises more slowly than the constant component. I.e. our left side will be larger than the right. Likewise if $\alpha_{k+1}<1$ the log component will shrink faster that the constant component which means it will also be less than the constant component. Therefore given our assumptions:

$$\ln \alpha_{k+1} \not\ge \alpha_{k+1}-1$$

and so:

$$P_{k+1}\left(\ln \alpha_{k+1} - \alpha_{k+1}\right) - P \not\ge -P - P_{k+1} + \epsilon$$

So much for the $k+1$th case. What about $k=1$. This is trivial because any $\alpha_1$ that satisfies:

$$\sum_k U_k = P-\epsilon$$

must be less than or equal to $1$ and therefore our product of quotients:

$$\prod_k \left(\frac{P_k\alpha_k}{P_k}\right)^{P_kN}>1$$

must be less than or equal to one.

\end{proof}

With this proof in hand we can now show how in the limit as the number of samples taken $N$ goes to infinity our likelihood is only maximized if $\hat{F} \rightarrow F$. \newline

For some given information $\theta$ and outcomes $V=\lbrace v_k \rbrace$ we have, in gathering our data, observed $v_k$ $N_k$ times. Therefore our overall likelihood is:

$$\mathcal{L} = \prod_k \hat{F}(v_k, \theta)^{N_k}$$

Now suppose that we represent:

$$\hat{F}(v_k, \theta)=F(v_k, \theta)\alpha_k$$

That is the ratio of our likelihoods between $\hat{F}$ and $F$ is given by:

$$\prod_k \left(\frac{F(v_k, \theta)\alpha_k}{F(v_k, \theta)}\right)^{N_k}$$

but we also know that:

$$\sum_k F(v_k, \theta) = \sum_k F(v_k, \theta)\alpha_k = 1$$

and that $\lim_{N\rightarrow \inf}N_k = F(v_k, \theta)N$ so that we have:

$$\prod_k \left(\frac{F(v_k, \theta)\alpha_k}{F(v_k, \theta)}\right)^{F(v_k, \theta)N}$$

which is now in the exact same form as our theorem above. And we now know that this ratio is maximized when $\alpha_k\equiv 1$. So as $N\rightarrow \inf$ a higher $\mathcal{L}$ means we are closer to approximating $F$.

>>>>

\section{Summary}

Vulnerability of fish stocks to various human and environmental impacts is tied to the spatio-temporal distribution and movement of those stocks. Therefore understanding these patterns is important to improving management practice today. Central to understanding these patterns is the creation of behavior models that can be used to test movement hypotheses and predict future and current movement behavior. Yet, as of today, there is no common methodology for building and using such models. The purpose of this study is to provide such a common methodology that takes advantage of modern computing solutions such as machine learning and the cloud. It is hoped that by providing such a strategy that (1) the barrier to producing movement models will be greatly reduced and (2) a common language for discussing, evaluating, and using these models will arise.

@@

\textbf{Target:} Discrete depth ranges. \newline

@@

\textbf{Target:} Spatial grid cell selection as movement. Also if clearly different modes of movement are found (migration vs foraging for example) we'll predict the mode as well. \newline

@@

\begin{enumerate}
\item Formatting Data for Odds Modeling 
\item Class Balancing
\item Splitting Data into Train, Validation, Test
\item Training and Hyperparameter Tuning Using Standard Neural Net Architectures
\item Diagnoses of Fit and Comparison of Different Model Types
\item Identifying Sample Groups that are More or Less Fit
\end{enumerate}

@@

\begin{figure}[h!] 
  \includegraphics[width=\linewidth]{simulation.png}
  \caption{Simulation Outline}
  \medskip
	\small
	Boxes with rounded edges would be inputs specified by the user. The hexagons are functions 		provided by the simulation and the trapezoids are data generated during the main loop. It is important to note that this design is anticipatory, not all mortality or environmental data may be available at this time in which case spatially homogeneous inputs would be provided.  
  \label{fig:simulation}
\end{figure}

@@

\begin{enumerate}
\item \textbf{Understanding Connectivity:} by looking at the vector flows of fish throughout the space we will be able to identify which areas are highly connected and which are more or less separated by environment, cohort, or other factors of fish behavior.
\item \textbf{Understanding Vulnerability:} by especially looking at changes in depth occupancy and preference for habitat we will attempt to quantify the degree of vulnerability to different kinds of activities as a function of space and time. Specific vulnerabilities of interest might include: vulnerability to specific fishing fleets, vulnerability to naval training activities, vulnerability to marine mammals. 
\item \textbf{Cumulative Mortality:} by being able to follow cohorts through space and time we should be able to show how different fishing pressure scenarios might affect the cumulative fishing mortality. Note that the data for specific fleets may be unavailable but that our interest here is to demonstrate that given data we can make these kinds of predictions.  
\item \textbf{Protected Areas/Times:} Of those fishing pressure scenarios, one of especial interest is modeling Marine Protected Areas (MPAs) or other kinds of closures in space and time. Given the motility of Chinook, how effective would marine protected areas be? 
\item \textbf{Uncertainty:} we will show that such a simulation can be run at scale, bootstrapping can be applied, and therefore uncertainty in simulation outcomes quantified. 
\end{enumerate}

@@

From studies that started with Beverton and Holt's work on recruitment in fisheries, we know, with a reasonable degree of certainty, that as a general rule fish stocks operate according to an asymptotic relationship between spawning biomass and recruits \citep{waltersmartell} (Fig. \ref{fig:recruitment}). What we see is a clear pattern in recruitment verses spawners for low spawners and then eventually we reach a point where this curve asymptotes and large changes to the number of spawners creates little to no change (especially when considering the actual variability in this data) in the number of recruits (there is some variation on this theme depending on the precise stock recruitment model used). Given the real variability in nature, this pattern looks more like a scatter shot of points but plenty of studies over several species have shown this overall pattern in the mean borne out \citep{waltersmartell}. \newline

@@

The implication for fisheries management is pretty clear - sustainable fishing is in large part driven by just maintaining the spawning stock biomass (SSB) above that knee in the curve \citep{waltersmartell} \citep{king}. Indeed this is well noted as the basis of why fisheries \textit{can} exist in the first place. If taking spawners always led to significant decreases in recruits, we'd quickly fish a stock out of existence. It is only through the compensation we get on the asymptote that sustainable fisheries are possible. However, two very simple observations of fish biology will show how actually managing SSB can be made difficult. \newline

The first, and simplest, is age structure. Obviously we're talking here about maintaining a specific level of spawning stock biomass. But today's juveniles are tomorrow's SSB and as fish grow they typically become more fecund. This means that in order to maintain our SSB over time we have to be careful about how we target each age group. This gets even more complicated for fish like Pacific salmon (\textit{Oncorhynchus}) or many species of groupers (\textit{Epinephelinae}) where life history is intimately entwined with spawning. The salmon only breed once \citep{NOAA24} and some species of groupers are protogynous hermaphrodites \citep{RedGrouper} meaning that if you target a specific age group you are effectively targeting a specific sex. Clearly understanding and managing our age class selectivity matters when it comes to maintaining a healthy SSB for years to come. \newline

The other observation is that many species of fishes actually form multiple sub stocks within what we may consider a single fishery (or sometimes many fisheries distributed across different countries) \citep{lorenzen2010} \citep{salmonplan} \citep{prince2010}. Pacific salmon are once again a good example in that they home to specific streams and only breed with other salmon in those streams and many other species of fish form fish spawning aggregations as well \citep{erisman2017}. Research also indicates that even fish like Atlantic herring likely form distinct genetic stocks due to differences in homing during breeding and school behavior \citep{armstrong2001}.

@@

All of this to say that understanding heterogeneity in selectivity by things like age class and sub stock is of the utmost importance for managing a fishery and truly ensuring an appropriate amount of SSB remains year over year. It is also worth noting that this knowledge around changes in selectivity in space and time is not only useful for considerations around SSB but also for other management issues such as bycatch avoidance and growth overfishing.

What kind of tooling do we need in order to address these issues and test out ideas for fisheries management? Clearly we need some kind of reliable simulation. \newline

@@

Clearly then this kind of data suits our modeling and simulation purposes well as we get behavior outcomes directly.

@@

Fisheries management is as diverse a subject as the species and contexts to which it is applied \citep{waltersmartell} \citep{king}. It simultaneously has to deal with creatures from Bluefin Tuna (\textit{Thunnus thynnus}) who make large oceanic migrations around the globe to various clams whose entire life cycles can be summarized within a few square meters \citep{prince2010}. There are species that school and species that consume those schools. Some live entirely at sea, some in inland rivers and lakes, and some, like salmon and eels spend time in both. There are species whose lives last the span of a year to others that can live longer than any us human beings.

And to capture this vast array of diversity are similarly diverse strategies and gears. From hooks to gillnets, from small boats to at sea factories, from recreational fishers to roving commercial fleets, and everything in between the variety in strategies that exist to catch fish are numerous \citep{waltersmartell} \citep{king}.

@@

First, it is clear that engagement is best facilitated by presentations of knowledge or data that are visual and relevant to the ways in which people normally think about their world \citep{prince2010} \citep{lorenzen2010}. By having simulations that can simply be shown and laid down on a map that places the spatio-temporal modeling in familiar context, one should be able to make objections or confirmation much easier. Rather than having to attempt to condense down loads of parameters, summary statistics, and the like in a model, one can simply show a map and folks can point at things that line up or are in disagreement with their intuition and experience. In this way mapped out views of the simulation work to be a useful device for consensus building among multiple groups of stakeholders. Not everyone may be familiar with the nuances of a specific kind of model but everyone with an interest can point to a map and understand what's going on. 
\newline

@@

For example, suppose that we have a more standard model that has been hand crafted by a scientist familiar with the species in question. They may have included specific associations with, say, oceanic currents, temperatures, and primary productivity. But then if someone looking at the simulation points out that they think the species' behavior is also being driven by day/night patterns the scientist then has to go back to the drawing board, figure out how to add this into their overall model, code it up, tune parameters, and then return - all without any real guarantee that whoever provided the feedback is right. And this would be for one piece of feedback. If there is a lot of feedback this only gets more difficult.

@@

Finally it is worth noting that the potential in combining this modeling with fishery dependent information is quite interesting on its own. Any forecasting model is always going to have issues predicting farther and farther into the future but fisheries dependent data could provide a means for making corrections to keep the model in line. This would amount to something like weather forecast where for the next few days the forecast is accurate enough to precisely predict specific kinds of movement helpful to fishers themselves.

@@

Fisheries assessments are one way in which unknown parameters important to management are estimated. By and large they are comprised of two models - (1) a state model that represents the underlying stock and contains such things as growth, age composition, recruitment, abundance, and so on and (2) an observation model that represents how data such as catch rates, fishery independent indices, and so on arise from the state model \citep{king}. Depending on how the assessment is being done the parameters in the state model will either be estimated independently or left as parameters in the model to be fit. The observation model on the other hand is encoded as a set of distributions dependent on those parameters (and including some parameters of its own) thus allowing for the likelihood of the parameters given the data to be computed \citep{sippel2014}. This, in turn, allows for maximum likelihood estimation (MLE) routines to then fit the likely parameters of the assessment. \newline

@@

In short the pattern is (1) design a simulation that has your unknown statistics as parameters, (2) specify how data is drawn from the simulation, (3) fit your parameters to your data using MLE. Then those desired statistics are extracted from the assessment (usually with some notion of uncertainty attached to them) and used to inform management.

@@

As has been noted by several authors, most standard assessments have typically assumed in their state model (or simulation) that the stock of interest is bounded by the area under assessment, is well mixed, and constitutes a single unit \citep{sippel2014} \citep{punt2019} \citep{goethel2023}. However from studies of fish behavior and the fact that fish obviously don't obey national or state defined boundaries, these assumptions are rarely true and indeed simulation studies have shown that assuming them can lead to significant errors and biases \citep{lorenzen2010}.

@@

Here then is another potential application of the kind of model proposed in this document. As we mentioned earlier these assessments are trying to model at least three things - recruitment, local population dynamics parameters such as growth or natural mortality, and then movement. By having a movement simulation based on real observations of the species in question we would be able to help build a high quality simulation of at least the last of those three. Rather than having to assume one form of movement or another we'd instead be able to see whether the assessment paradigm in question can capture the relevant aspects of structure in the \textit{real} movement of our stocks and sub-stocks. And this in turn would help in selecting the assessments best suited to the job (as well as illustrate what uncertainties exist if a lack of specific kinds of data requires us to use an over-simplified assessment paradigm).

@@

\begin{enumerate}
\item \textbf{High Precision as a Requirement} - IBMs typically predict what an animal \textit{will} do as opposed to the likelihoods of what it \textit{could} do. This creates the requirement for capturing \textit{all} the meaningful drivers of variability in the behavior of the animal. To see why this is consider the example of a creature that can either choose to stay put or move to another location. If the more likely of the two actions is to stay put a deterministic model \textit{must} predict that the animal will \textit{always} stay put. But this is obviously non-sensical. However, because the model is deterministic in nature the modeler cannot depend on stochasticity to save the day - instead they must find whatever is driving the difference between staying put and moving on. And this can cascade into greater and greater requirements for modeling the "internal state" of an animal. 
\item \textbf{Obscurity} - given the inherent complexity of IBMs that arises from the point above it gets increasingly difficult to make IBMs clear to new users or potential developers. There can often be many dozens of different parameters, functions, and "internal states" to keep track of without even considering all of the external modeling that has to go in place to provide for all of this. And this makes for not just complicated descriptions but equally if not more complicated code.
\item \textbf{Hand Made} - almost all IBMs have been hand-coded in the sense that every functional relationship is specifically crafted and outlined by the modeler. This is obviously a very laborious process but it also means that no two IBMs are the same and that lack of generalization or abstraction makes exchanging lessons learned near difficult to impossible. 
\item \textbf{No Generalization of Fitting or Diagnostics} - because each IBM is effectively unique the methods for fitting, the data used, and the diagnostics developed are all unique as well. This once again increases the workload of the modeler, makes exchanging lessons learned and best practices extremely difficult, and also makes starting a new model from scratch quite prohibitive. 
\end{enumerate}

@@

The immature fish then spend time (up to five years) growing in the marine environment before maturing and then migrating back to the river where they were born to breed and subsequently die \citep{oceanchinook}. The timing of such homing behavior is clearly season dependent with species further north having to avoid ice build up during the cooler months \citep{oceanchinook}. The mechanism behind this natal homing ability is still a question under active research \citep{bracis2012}. \newline

Given this homing behavior, Chinook salmon populations are in fact a series of reasonably separated genetic sub-populations each of which can have its own life history strategy. Indeed using tagging data it has been demonstrated that both the initial and later marine distributions of Chinook salmon is highly dependent on the region of origin for the fish in question \citep{shelton2019} \citep{tucker2019}.

@@

The initial purpose of these hatchery programs was to supplement losses in juveniles due to targeting of adult fish in river systems, however the programs have not created the kind of recovery that was intended and it is unclear how wild and hatchery populations interact. For example it is well known that there is quite a bit of genetic exchange between wild and hatchery types and concern that hatchery types are less suited to wild environments \citep{oceanchinook}. \newline

There are valuable tribal, commercial, and recreational fisheries of Chinook salmon throughout the Pacific \citep{salmonplan} yet given the highly mobile nature of Chinook salmon independent management by sovereign nations is simply not possible and so management is the ultimate responsibility of the Pacific Salmon Commission. Federal management councils (NOAA for the United States) then collaborate with the Pacific Salmon Commission to set harvest quotas and other management objectives. The governments of Cananda and the United States then work with states, tribes, and industry to create and manage shared objectives based on these overall goals \citep{NOAA22}.

@@
